---
title: "R for SME: practical 9 - Logistic regression 1"
author: Andrea Mazzella
output: html_document
---
> 0 Basics

Install these packages if you haven't already
```{r}
install.packages("readstata13") 
install.packages("epiDisplay") 
install.packages("magrittr")
install.packages("tidyr")
install.packages("dplyr")
```

Load packages
```{r}
library(readstata13)
library(epiDisplay)
library(magrittr)
library(tidyr)
library(dplyr)
```
> 1

Make sure you have the mortality.dta dataset in the same folder as this .rmd

> 2

Import data
```{r}
mortality<-read.dta13("./mortality.dta", nonint.factors = T)
```

Explore data
```{r}
head(mortality)
glimpse(mortality)
summary(mortality)
View(mortality)
```

> 3

Outcome variable: died (3-year-mortality; binary, 0/1)
Exposure variable: vimp (visual impairment; binary, "Normal" / "Visually impaired")
- Examine the association between visual impairment and death
```{r}
# Frequency table with %
mortality %$% tabpct(vimp, died, percent = "row", graph = F)

# OR
mortality %$% cc(vimp, died, graph = F)
```

> 4

Perform a logistic regression to examine the association between visual impairment and death.
```{r}
# Logistic regression of a binary exposure

logit_vimp <- glm(died ~ vimp, data = mortality, family = binomial()) # create the model

logistic.display(logit_vimp) # get a more readable output
```
Output:
- OR with 95%CI for outcome in exposed vs unexposed, Wald's test and LR-test p-value
- (final) log-likelihood - (no intermediate iterations shown in R, unlike in stata)
- number of observations
- AIC value

> 6

Unlike Stata, R already gives you the answers in OR - no need to show useless log(OR).

> 8

Explore the association between microfilaria and death with crosstabulation:
- Exposure: microfilarial load/mg - mfgrp (categorical variable: Uninfected, <10, 10-49, 50, NA)
```{r}
mortality %$% summary(mfgrp)

mortality %$% tabpct(mfgrp, died, percent = "row", graph = F)
```


> 9

Now explore the same with logistic regression.
Unlike Stata, R understands that this variable is categorical, so the formula is the same as for a binary variable.

```{r}
logit_mfgrp <- glm(died ~ mfgrp,
                 data = mortality,
                 family = binomial())
logistic.display(logit_mfgrp)
```

> 13

To perform a likelihood ratio test (LRT), you need to use lrtest() to divide the log likelihood from a logistic regression model *with* the variable (that you have already defined) and the log likelihood from a logistic regression model *without* the variable - that you need to calculate now.

Caution: mfgrp has missing data, so this new model will have more observations than the first - and the LRT test can only work when the two models have the same number of observations. So you need to 
```{r}
# Check for missing values
mortality %$% summary(mfgrp)

# Logistic regression model without the covariate (and removing observations with a missing value in mfgrb)
logit_0 <- mortality %>%
  drop_na(mfgrp) %>%
  glm(died ~ 1,
      data = .,
      family = binomial())

# Likelihood ratio test (LRT)
lrtest(logit_0,logit_mfgrp)
```

> 14

Run a logistic regression model to check the association of age and death.
Exposure variable: agegrp (categorical: 15-34, 35-54, 55-64, 65+)
```{r}
# Check for missing values
mortality %$% summary(agegrp)

# Logistic regression model for age
logit_age <- mortality %>%
  glm(died ~ agegrp,
      data = .,
      family = binomial())
logistic.display(logit_age)
```

How do we use two exposure variables in a logistic regression model?
- Use both visual impairment (vimp) and age (agegrp)
```{r}
# Logistic regression model with 2+ exposures
logit_vimp_age <- mortality %>%
  glm(died ~ vimp + agegrp,
      data = .,
      family = binomial())
logistic.display(logit_vimp_age)
```
