---
title: "R for SME 12: Logistic regression (quantitative exposures)"
author: Andrea Mazzella [link](https://github.com/andreamazzella)
output: html_notebook
---

## Basics
Load packages
```{r Load packages}
library(haven)
library(pubh)
library(rstatix)
library(epiDisplay)
library(magrittr)
library(tidyverse)
```

# Part 1: mortality

## Data import, exploration, management

1.

Make sure you have the mortality.dta dataset in the same folder as this .rmd

```{r Import data}
mortality<-read_dta("./mortality.dta")
mortality %<>% mutate_if(is.labelled,as_factor)
```

1a.

Explore the data.
What type of variables are "systolic" and "died"?
```{r Explore data}
glimpse(mortality)
summary(mortality)
View(mortality)
```
- systolic is a quantitative variable
- died is a binary variable coded as 0/1, but its class is "double" and not "factor".

```{r Transform "died" into a factor variable}
mortality %<>%
  mutate(died = factor(
    died,
    levels = c(0, 1),
    labels = c("alive", "dead")
  ))
```


Plot systolic BP against death. What does the plot show?
```{r Visualising}
# Scatterplot
mortality %>% ggplot(aes(x = systolic, y = died)) +
  geom_point() +
  theme_bw() +
  labs(title = "Systolic blood pressure and death",
       x = "Systolic blood pressure",
       y = "Death")
## I am not sure why that's what they wanted us to do in the STATA practical. A scatteplot is only helpful with two continuous variables. With 1 continuous and 1 categorical, you use Box plots, which can give you a vague idea of how did sBP compare in dead vs alive.

# Box-and-whiskers plot
mortality %>% ggplot(aes(x = died, y = systolic)) +
  geom_boxplot() +
  labs(title = "Systolic blood pressure and death",
       y = "Systolic blood pressure",
       x = "Death") +
  theme_bw() +
  coord_flip()
```

1b.

Group systolic into three levels and add labels.
- <120: normal
- 120-139: pre-hypertension
- 140+: hypertension
```{r Categorise "systolic"}
mortality %<>%
  mutate(systolic_grp = cut(
    systolic,
    breaks = c(0, 119, 139, +Inf),
    labels = c("normal", "pre-hypertension", "hypertension")
  ))

# Check it worked ok
mortality %$% table(systolic, systolic_grp)
```

Explore the relation between systolic_grp and died.
```{r Crosstab systolic_grp and died}
mortality %$% tabpct(systolic_grp, died, percent = "row", graph = F)
```
- A higher percentage of hypertensives died than people with normal systolic BP.

1c, 1d.

Calculate the log{odds} of death in each sBP group, and plot them manually on a graph, then draw a line by eye to fit these points.
_Issue:_ I don't know how to get odds in R.
_Workaround:_ odds_trend() gives you the OR for each group of systolic BP and the baseline (normal systolic BP):

```{r OR}
# OR
odds_trend(died ~ systolic_grp, data = mortality)

# Test for trend
death_by_BP <- mortality %$% table(died, systolic_grp)
prop_trend_test(death_by_BP)
```
- pre-hypertension: OR 1.32 (0.87-2.00)
- hypertension: OR 3.63 (2.38-5.52)
- this command also gives you a graph (but not on the log scale)
- chi2 test for trend: p <0.001

1e.

Fit a logistic regression model for death, with exposure being grouped sBP treated as a quantitative variable. In order to do this, you simply need to transform the variable back to a numeric variable, with values of 1, 2, 3. Once you do that, the glm() command is the same.

```{r Logistic regression with a linear relation}
# Recode systolic_grp as numeric
mortality %<>%
  mutate(systolic_quant = as.numeric(systolic_grp))

# Check
mortality %$% table(systolic_grp, systolic_quant)

# Logistic regression with linear relation - log odds scale
logit_linear <- glm(died ~ systolic_quant,
                    data = mortality,
                    family = binomial())
logit_linear
```
- What is the interpretation of the estimates for grouped sBP?


1f.

Now do the same in the odds (ie: non-log) scale.
```{r}
# Logistic regression with linear relation - odds scale display
logistic.display(logit_linear)
```
- What is the interpretation of the estimates for grouped sBP now?

1g.

Fit a model with grouped sBP assuming factors and perform a LRT on the following hypotheses:
- H0: the effect of grouped sBP on log(odds) of death is linear;
- H1: the relationship is not linear.
```{r}
logit_nonlin <- glm(died ~ systolic_grp,
                    data = mortality,
                    family = binomial())
logistic.display(logit_nonlin)

# LRT
lrtest(logit_linear, logit_nonlin)
```

1h.

Summarise your findings
- There is very strong evidence for a linear association between grouped sBP and death; OR = 1.86 (1.49,2.32)  Wald's p < 0.001 
- There is only weak evidence for a non-linear relationship between systolic BP and death (LRT p = 0.06)

1i.

Recode the sBP groups so that it's a quantitative variable, and each group contains a mid-point value in mmHg of that group.
Reminder of systolic BP categories:
- <120: normal
- 120-139: pre-hypertension
- 140+: hypertension
```{r}
# Recode the variable to numerical with (arbitrary) midpoints
mortality %<>%
  mutate(systolic_grp_mid = as.numeric(
    case_when(
      systolic_grp == "normal" ~ 110,
      systolic_grp == "pre-hypertension" ~ 130,
      systolic_grp == "hypertension" ~ 150
    )
  ))

# Check you haven't made a mess
mortality %$% table(systolic_grp, systolic_grp_mid)
```

Now fit the equivalent model to 1f.
```{r}
glm(died ~ systolic_grp_mid,
    data = mortality,
    family = binomial()) %>%
  logistic.display()
```
- What is the interpretation of the estimate for grouped sBP now?

2a.

Consider a model with
- outcome: death;
- exposure: visual impairment;
- confounder: age (agegrp)
Does our estimate of effect of visual impairment differ if we treat age as quantitative rather than categorical ?
Fit a linear and a non-linear model.

```{r}
mortality %$% class(agegrp)

# Transform agegrp into a quantitative variable
mortality %<>%
  mutate(agegrp_lin = as.numeric(agegrp))

mortality %$% table(agegrp, agegrp_lin)
mortality %$% class(agegrp_lin)

# Quantitative model
logit_linear2 <- glm(died ~ vimp + agegrp_lin,
                     data = mortality,
                     family = binomial())
logistic.display(logit_linear2)

# Categorical model
logit_nonlin2 <- glm(died ~ vimp + agegrp,
                     data = mortality,
                     family = binomial())
logistic.display(logit_nonlin2)

# LRT
lrtest(logit_linear2, logit_nonlin2)
```
- The adjusted OR are very similar in the two models
- LRT p-value = 0.786
- There is no evidence that for non-linearity.


2b.

Perform a LRT for the interaction between vimp and agegrp treated as a linear effect.
Is there evidence that the association of visual impairment on death differs by age group, where age group is treated as a linear effect?
```{r}
mortality %<>%
  mutate(
    agegrp2 =
      case_when(
        agegrp == "15-34" ~ 0,
        agegrp == "35-54" ~ 1,
        agegrp == "55-64" ~ 2,
        agegrp == "65+" ~ 3
      )
  )
mortality %$% table(agegrp2, agegrp)

# Logistic regression with interaction
glm(died ~ vimp * agegrp2,
    data = mortality,
    family = binomial()) %>%
  logistic.display()

# LRT


# Calculate stratum-specific ORs for visual impairment for the four age group strata



```

3.

Using the Mwanza dataset (MWANZA.dta), investigate the association between HIV infection and number of injections in the past year.

```{r}
# Import the dataset
mwanza <- read_dta("./mwanza.dta")
```

```{r}
#Familiarise yourself with the data
View(mwanza)
glimpse(mwanza)
summary(mwanza)
```

The variable "inj" is coded as follows:
1: does not inject
2-4: increasing number of injections per year
9: missing value
```{r}
mwanza %<>%
  mutate(hiv = as.factor(case_when(case == 1 ~ "HIV+",
                                   case == 0 ~ "negative")))

# Replace inj "9" with "NA"
mwanza$inj <- na_if(mwanza$inj, 9)
table(mwanza$inj)
mwanza %$% tabpct(hiv, inj, graph = F)

# Logistic regression including the zero group
glm(hiv ~ inj,
    data = mwanza,
    family = binomial()) %>%
  logistic.display()

```

Remember that the zero group should be excluded in order to confirm that a trend with increasing numbers is not simply induced by a difference between the zero category and the rest.

```{r}
# Logistic regression excluding the zero group
mwanza %>%
  filter(inj != 1) %>% # this line temporarily removes all observations with a value of `1` in "inj"
  glm(hiv ~ inj,
      data = .,
      family = binomial()) %>%
  logistic.display()
```
















