---
title: "R for SME 13: Matched case-control studies"
author: Andrea Mazzella [link](https://github.com/andreamazzella)
output: html_notebook
---

## Basics
```{r Load packages}
library(haven)
library(epiDisplay)
library(magrittr)
library(survival) # for conditional regression
library(dplyr)
```

#Part 1: individual matching (paired data)

## 1: Data import and exploration

Make sure you have the diabraz.dta and diabraz2 datasets in the same folder as this .rmd file.
```{r - Import}
diabraz <- read_dta("./diabraz.dta")
diabraz2 <- read_dta("./diabraz2.dta")
```

Explore the data in diabraz. We're interested in variables "bf" and "bwtgp".
- How many infants do you have data for?
```{r - Explore}
View(diabraz)
glimpse(diabraz)
summary(diabraz)
```
- Data management nightmare - they're all doubles but they contain integers which code for either binary/categorical or continuous variables. Aaaand there are no labels. *Thanks, diabraz!*
  From the instructions:
  - "case" = "death secondary to diarrhoea"
  - "bf" = breastfed (yes, no)
  - "bwtgp" = birthweight group (>=3kg, <3kg)
We won't recode these now because we will need them in a special format for a function below.

- 172 observations

- How many cases (deaths from diarrhoea) and how many controls?
- What does "set" refer to?
```{r}
diabraz %$% table(case)
diabraz %$% table(case, set)
```
- 86 cases and 86 controls
- each "set" consists of a pair of 1 case and 1 control

## 2, 3: Matched analysis for breastfeeding

Now let's create a matched table, which is a 2x2 table with cases in the columns, controls in the rows. Each cell will contain the number of pairs with a given combination of exposure and outcome.

The function is epiDisplay::matchTab(); it takes as arguments, in this order:
- the case variable,
- the binary exposure variable (coded as 0/1),
- the pair variable

```{r Matched table}
# Recode bf variable
diabraz %<>%
  mutate(breastfed = as.factor(case_when(bf == 1 ~ 0,
                                         bf == 2 ~ 1)))
diabraz %$% table(bf, breastfed)

# Create matched table and get OR
diabraz %$% matchTab(case, breastfed, set)

# chi2 test
## brace yourself for a huge (mostly useless) table
diabraz %$% epiDisplay::mhor(case, breastfed, set, graph = F)
```
- There are 24 pairs in which both the case and the control were not breastfed,
- 6 pairs in which the control was breastfed but not the case.
- 29 pairs in which the case was breastfed but not the contorl,
- 27 pairs in which neither case nor control was breastfed
NB: unlike Stata, this R command also calculates the OR:
- OR by MH method = 4.83, 95%CI 2.01 - 11.64
- chi2 = 15.1, p < 0.001

## 4: Matched analysis for birthweight

Now do the same for birthweight (bwtgp) and diarrhoea mortality.
```{r Matched table}
# Recode bwtgp variable
diabraz %<>%
  mutate(birthweight = as.factor(case_when(bwtgp == 1 ~ 0,
                                           bwtgp == 2 ~ 1)))
diabraz %$% table(bwtgp, birthweight)

# Create matched table and get OR
diabraz %$% matchTab(case, birthweight, set)

# chi2 test
diabraz %$% epiDisplay::mhor(case, birthweight, set, graph = F)
```
- OR by MH = 1.39 (0.76 - 2.55)
- chi2 = 1.14, p = 0.29

## 5: What if we did an unmatched analysis?

Construct unmatched tables and calculate unmatched estimates of the ORs with cc().
What is the effect of ignoring matching?
Why is it not the same in both exposures?

```{r}
diabraz %$% cc(case, breastfed, graph = F)
diabraz %$% cc(case, birthweight, graph = F)
```
- It gives a different estimate of the OR


## 6: Conditional regression

Let's analyse the association between breastfeeding and diarrhoeal mortality (from question 2/3) with conditional regression.
The function is clogit() from {survival}. It's very similar to glm; you only need to specify the pair variable in strata(); there's no need to specify the family.

```{r}
clogit(case ~ breastfed + strata(set),
       data = diabraz) %>%
  clogistic.display()
```
- OR 4.83 (2.01,11.64): same result as in question 2/3.
- Wald's test p < 0.001

And now let's do the same for birthweight (same as question 4 but with conditional regression).
```{r}
clogit(case ~ birthweight + strata(set),
       data = diabraz) %>%
  clogistic.display()
```
- Again, equivalent OR results.

## 7: A more complex dataset

Explore the data in diabraz2, which has the same variables as diabraz.
- How many cases and how many controls?
```{r}
diabraz2 %$% table(case)
diabraz2 %$% table(case, set)
```
- There are 2 controls for each case.

Create a matched table. In this table, each cell refers to a group of 1 case and its 2 controls.
For some reason in this dataset breastfeeding has been coded the other way round: 1 means breastfed, 2 means not breastfed. So we need to swap these. *Thanks, diabraz2!*
```{r Matched table}
# Recode values of bf
diabraz2 %<>% mutate(breastfed = as.factor(case_when(bf == 1 ~ 1,
                                                     bf == 2 ~ 0)))
diabraz2 %$% table(bf, breastfed)

# Create matched table and get OR
diabraz2 %$% matchTab(case, breastfed, set)

# chi2 test
diabraz2 %$% epiDisplay::mhor(case, breastfed, set, graph = F)
```
- Interpret all the cells' values.
- Calculate how many total cases. *SUM(table)?*
- How many cases are breastfed? *21+24+5?*
- OR-MH = 4.06
- OR (MLE) = 3.79 (2.47-5.82)
- chi2 p-value < 0.001

Let's do the same analysis with conditional regression:
```{r}
clogit(case ~ breastfed + strata(set),
       data = diabraz2) %>%
  clogistic.display()
```
- OR is similar to MH.
- Wald's test p < 0.001

## 13: Conditional regression with confounder

Investigate whether social class (social) or maternal education (meduc) confound the association between breastfeeding and diarrhoeal mortality.
```{r}
# Convert social and meduc into factors
diabraz2 %<>% mutate(social2 = as.factor(social))
diabraz2 %<>% mutate(meduc2 = as.factor(meduc))

# Logistic regression with confounding covariates
clogit(case ~ breastfed + social2 + meduc2 + strata(set),
       data = diabraz2) %>%
  clogistic.display()
```
- There is still very strong evidence for association between breastfeeding and diarrhoeal mortality even after adjusting for social class and maternal education (p < 0.001); actually, there is a slightly stronger association: OR went from 3.79 (crude) to 4.28 (adjusted)

*Part 2: frequency matching*
(Pen and paper only)