---
title: "R for SME: Logistic regression 3 (practical 11)"
author: Andrea Mazzella
output: html_document
---

```{r - Load packages}
library(readstata13)
library(epiDisplay)
library(magrittr)
library(tidyr)
library(dplyr)
options(digits = 3, scipen = 999) # Round all output to 3 decimal points and remove scientific notation
```

*Part 1: mortality*

> 1

Make sure you have the mortality.dta dataset in the same folder as this .rmd file.
```{r - Import}
mortality <- read.dta13("./mortality.dta", nonint.factors = T)
```

```{r - Explore}
head(mortality)
glimpse(mortality)
summary(mortality)
View(mortality)
```

Analyse the association between visual impairment and death, stratifying on sex, using a Mantel-Haenszel approach.
```{r}
# 2x2 table
mortality %$% tabpct(died, vimp, percent = "col", graph = F)

# Crude OR
mortality %$% cc(died, vimp, graph = F)

# Stratified tables
mortality %$% table(died, vimp, sex)

# Stratum-specific OR and MH-OR
mortality %$% mhor(died, vimp, sex, graph = F)
```
Crude OR: 5.57 (3.78-8.2)

Stratum-specific OR (CI)
- male   3.94       2.15       6.98
- female 8.05       4.41      14.29

MH OR  5.43       3.69       8.00

Homogeneity test p = 0.07: there is weak evidence of interaction between sex and visual impairment; the OR in females is about twice the OR in males.

> 2

Fit logistic regression models to estimate the same association, without interaction and with an interaction.
```{r}
# Model without interaction
glm(died ~ vimp + sex,
    data = mortality,
    family = binomial()) %>%
  logistic.display()

# Model with interaction
glm(died ~ vimp * sex, # the asterisk marks the interaction
    data = mortality,
    family = binomial()) %>%
  logistic.display()
```
Model without interaction:
Adj OR (died/vimp//sex) = 5.53 (3.75,8.16)  (Wald's p < 0.001)

Model with interaction:
Adj OR (died/vimp//sex) = 3.95 (2.28,6.84)  (Wald's p < 0.001), vimp+female 2.04 (0.94,4.44)

*issue* the glm output doesn't show the equivalent of STATA's "_cons", so you can't calculate the table for OR in the four groups?

> 3

...

> 4

Fit a logistic regression model with interaction between visual impairment and sex, and control for age
```{r}
glm(died ~ vimp * sex + agegrp,
    data = mortality,
    family = binomial()) %>%
  logistic.display()
```
- OR for visual impairment in males: ???
- OR for females in visually unimpaired: 0.85
- Interaction parameter between vimp and sex: 1.93
- OR for visual impairment in females: ??
- Impact on controlling for age: 

> 5

log(OR) - don't know how to get this.

*Part 2: Mwanza*

> 6

Make sure you have the mwanza.dta dataset in the same folder as this .rmd file.
```{r}
# Import data
mwanza <- read.dta13("./mwanza.dta", nonint.factors = T)
glimpse(mwanza)
```

Use logistic regression to assess whether the association between education (ed) and HIV (case) is modified by age (age1). First, relevel these variables:
- ed into two groups: "none" and "any formal education";
- age1 into three groups: 15-24, 25-34, 34+.
```{r}
# Relevel "ed" and "age1"
mwanza <- mwanza %>%
  mutate(ed2 = as.factor(
    case_when(
      ed == 1 ~ "none",
      ed == 2 ~ "any formal",
      ed == 3 ~ "any formal",
      ed == 4 ~ "any formal"
    )
  ))
mwanza %$% table(ed,ed2)

mwanza <-
  mwanza %>% mutate(age3 = as.factor(
    case_when(
      age1 == 1 ~ "15-24",
      age1 == 2 ~ "15-24",
      age1 == 3 ~ "25-34",
      age1 == 4 ~ "25-34",
      age1 == 5 ~ "34+",
      age1 == 6 ~ "34+",
    )
  ))
mwanza %$% table(age1,age3)
```

```{r}
# Crude OR
mwanza %$% cc(case, ed2, graph = F)

# Stratum-specific OR and MH-OR
mwanza %$% mhor(case, ed2, age3, graph = F)

# Logistic regression without interaction
logit_without <- glm(case ~ ed2 + age3,
                     data = mwanza,
                     family = binomial())

# Logistic regression with interaction
logit_interact <- glm(case ~ ed2 * age3,
                      data = mwanza,
                      family = binomial())

lrtest(logit_without, logit_interact)
```


Crude OR: 0.41 (0.29-0.60) p < 0.001
MH OR:    0.43 (0.29-0.63) p < 0.001
Homogeneity test, p-value = 0.006. There is strong evidence for interaction between age and education.
Interaction parameter: 0.2 in age group 25-34, 0.41 in age group 34+.

- Is there evidence of interaction between age and education?































